## Code produced to run through peer review assignment 1
##### Written by Dan Kellett
##### 10/03/2015

```{r}
setwd("C:/Users/Dan & Rach/Desktop/Data Science/Reproducible Research")
```

## Loading and preprocessing the data

##### Load the data (i.e. read.csv())
##### Data loaded into data frame 'rawData' by unzipping the file directly from GitHub
```{r}
rawData <- read.table(unz("C:/Users/Dan & Rach/Documents/GitHub/RepData_PeerAssessment1/activity.zip"
                          ,"activity.csv"), sep = ",", header = TRUE)
```

##### Process/transform the data (if necessary) into a format suitable for your analysis
##### Convert the 'date' field into a date format
```{r}
rawData$date <- as.Date(rawData$date, "%Y-%m-%d")
```

## What is mean total number of steps taken per day?
##### For this part of the assignment, you can ignore the missing values in the dataset.

##### Calculate the total number of steps taken per day
##### Use the doBy package to sum steps data by date - put results into data frame 'stepsSum'
```{r}
library(doBy)
stepsSum <- summaryBy(steps ~ date, data=rawData, FUN=c(sum))
stepsSum
```

##### If you do not understand the difference between a histogram and a barplot, 
##### research the difference between them. Make a histogram of the total number of steps 
##### taken each day
##### Use base plot to create a histogram with some basic re-labelling
```{r}
hist(stepsSum$steps.sum, xlab="Total steps by day", main="Histogram of total steps by day")
```

##### Calculate and report the mean and median of the total number of steps taken per day
##### Use mean and median functions to calculate
```{r}
mean(stepsSum$steps.sum, na.rm=TRUE)
median(stepsSum$steps.sum, na.rm=TRUE)
```

## What is the average daily activity pattern?

#### Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and 
#### the average number of steps taken, averaged across all days (y-axis)
#### Use the doBy package to average steps data by interval - put results into data frame 'intervalAve'
#### Then use base plot to create a labelled time series plot
```{r}
intervalAve <- summaryBy(steps ~ interval, data=rawData, FUN=c(mean), na.rm=TRUE)
plot(intervalAve$interval, intervalAve$steps.mean, xlab="5-minute interval", ylab="Average steps", 
     type="l", main="Time series plot of average steps by 5-minute time interval")
```

#### Which 5-minute interval, on average across all the days in the dataset, 
#### contains the maximum number of steps?
#### Create a value 'maxStep' with the maximum average step value then subset intervalAve by this
```{r}
maxStep <- max(intervalAve$steps.mean)
maxInterval <- subset(intervalAve, steps.mean == maxStep)
maxInterval
```

## Imputing missing values
##### Note that there are a number of days/intervals where there are missing values (coded as NA). 
##### The presence of missing days may introduce bias into some calculations or summaries of the data.

##### Calculate and report the total number of missing values in the dataset (i.e. the total number 
##### of rows with NAs)
##### Use sum function to count number of rows where steps = NA
```{r}
sum(is.na(rawData$steps))
```

##### Devise a strategy for filling in all of the missing values in the dataset. The strategy does 
##### not need to be sophisticated. For example, you could use the mean/median for that day, or 
##### the mean for that 5-minute interval, etc.
##### Merge raw data with data containing average #steps by interval
```{r}
names(intervalAve)[2] <- "activityLabels"
mergeData <- merge(rawData, intervalAve)
```

##### Create a new dataset that is equal to the original dataset but with the missing data filled in.
##### New mergeData data frame contains data with NA replaced by average steps by interval
```{r}
mergeData$steps[is.na(mergeData$steps)] <- mergeData$activityLabels[is.na(mergeData$steps)]
mergeData$activityLabels <- NULL
```

##### Make a histogram of the total number of steps taken each day and Calculate and report the 
##### mean and median total number of steps taken per day. Do these values differ from the estimates 
##### from the first part of the assignment? What is the impact of imputing missing data on the 
##### estimates of the total daily number of steps?
```{r}
stepsSum2 <- summaryBy(steps ~ date, data=mergeData, FUN=c(sum))
stepsSum2
hist(stepsSum2$steps.sum, xlab="Total steps by day", 
     main="Histogram of total steps by day (missing data imputed)")
mean(stepsSum2$steps.sum, na.rm=TRUE)
median(stepsSum2$steps.sum, na.rm=TRUE)
```
##### Imputing data increases the total number of steps per day but the mean remains unchanged

## Are there differences in activity patterns between weekdays and weekends?
##### For this part the weekdays() function may be of some help here. Use the dataset with 
##### the filled-in missing values for this part.

##### Create a new factor variable in the dataset with two levels - "weekday" and "weekend" 
##### indicating whether a given date is a weekday or weekend day.
##### Create a new column (IsWeekend) based on a logical assessment of the day using the weekdays() 
##### function (TRUE if Saturday or Sunday, else FALSE)
```{r}
mergeData$IsWeekend <- weekdays(mergeData$date) %in% c('Sunday','Saturday')
```

##### Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval 
##### (x-axis) and the average number of steps taken, averaged across all weekday days or weekend 
##### days (y-axis). See the README file in the GitHub repository to see an example of what this 
##### plot should look like using simulated data.
##### Summarise the data using summaryBy function then panel plot using ggplot2 library
```{r}
intervalAve2 <- summaryBy(steps ~ interval + IsWeekend, data=mergeData, FUN=c(mean), na.rm=TRUE)
library(ggplot2)

g <- qplot(interval, steps.mean, data=intervalAve2, facets=IsWeekend~., ylab="Mean number of steps",
           main="Average steps by interval: 
      Weekend (TRUE) vs. weekday (FALSE)")
g + geom_line() + facet_grid(IsWeekend~., scales = "free")
```
